PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-10-17 IEEEtran/paper.tex
Final edit before minor rewiews

% \ifANNOTATED
\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\DeclareRobustCommand{\add}[1]{\textcolor{ao(english)}{#1}}%\sethlcolor{lightgreen}\hl{#1}}
\DeclareRobustCommand{\remove}[1]{}
\DeclareRobustCommand{\addCite}[1]{\textcolor{lightgreen}{\mbox{#1}}}
\newcommand{\TG}[1]{\color{red}\textsc{From Tristan}: #1\xspace\color{black}}
% \else
% \DeclareRobustCommand{\add}[1]{#1}
% \DeclareRobustCommand{\remove}[1]{}
% \DeclareRobustCommand{\addCite}[1]{#1}
% \newcommand{\TG}[1]{}
% \fi

@@

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (\add{a}\remove{1}) generating numerically-perturbed
executions, (\add{b}\remove{2}) tracing floating-point computations along with
program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. \add{PyTracer addresses (a) by relying on
    Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
    computing libraries instrumented with Monte-Carlo arithmetic through the
    Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
    Python functions, modules, and classes through meta-programming, and (c) by
    visualizing stability measures throughout program executions with an
    interactive Plotly dashboard. \mbox{\pytracer} is available
    \mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements
    with (1) a Fuzzy Python interpreter instrumented with Monte-Carlo
    arithmetic, (2) a dynamic instrumentation of Python functions, modules, and
    classes through meta-programming, (3) an interactive Plotly dashboard to
    visualize stability measures throughout program executions. }

@@

\add{
    To detect numerical instabilities, \pytracer works in three main steps:
    \begin{enumerate}
        \item It instruments Python programs and their module dependencies
              automatically, to save the input and output values of functions in a trace
              file.
        \item It executes the instrumented program several times using Monte-Carlo
              Arithmetic to inject perturbations in function outputs.
        \item It aggregates all trace files to compute statistics about the traced
              values, and visualize them in a GUI.
    \end{enumerate}
    \pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
    In step 1, \pytracer uses monkey patching to replace the functions called
    by the application with a wrapper that saves their arguments and returned
    values in a trace file. In step 2, it executes the application multiple
    times with Monte-Carlo Arithmetic as implemented in the
    Fuzzy~\cite{kiar2020comparing} environment. It then
    aggregates the resulting trace files to compute summary statistics on
    traced variables. In step 3, a Plotly dashboard provides interactive
    visualizations highlighting unstable code sections.
}

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis \TG{[23] fix ref}, and random
seed analysis \TG{fix ref [11]}.  \add{PyTracer is agnostic to the noise model used and can therefore
    be used for all these methods}.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis\add{.}

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
    simulates roundoff errors by applying function $inexact$ only to the
    output returned by a function. The Precision Bounding (PB) mode can
    identify other types of instabilities, in particular catastrophic
    cancellations, however, it is more difficult to apply to large code bases
    as it often leads to execution errors that require human intervention.}

@@

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate the
    distribution will be. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample sizes (3 to 30)
    can give good estimates, the issue is to capture ``rare events" that could make the
    computation strongly diverge from the mean. Determining the number of
    MCA samples to use for a given instrumented code is still an open research question. 
    Recent theoretical works}~\cite{sohier2021confidence}
\add{on the estimation of significant digits allow computing an associated
    confidence interval depending on the sample size and the underlying distribution.}

@@

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
    PyTracer. We observe a significant loss of \add{result} precision in some regions, which follows
    an interesting pattern that could be further investigated.}

@@

Replacing \texttt{gesdd} for \texttt{gesvd} solved the SVD convergence issue.
\pytracer allows to quickly visualize the new \add{result} precision.
Figure~\ref{fig:face_recognition_svd} shows the \add{result} precision of the eigenvalues
computed with \texttt{gesvd} \add{in single precision}. \add{The result precision}
varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. \add{We can note that even with the \texttt{gesvd} method,
    the \add{result} precision of the singular values and eigenvectors is not uniform. 
    Indeed, we can see that the \add{result} precision of the lowest singular values and eigenvectors decreases 
    until reaching 0 significant bits.}

@@

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues \TG{(such as ...)},
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request has recently been submitted \TG{in the ... repo} to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unified view of all traces allows the user to quickly navigate through the
    \add{result} precision of different variables.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}

@@

\add{\pytracer can be used with different types of numerical perturbations,
    including random seeds, data noise, or stochastic arithmetic. The
    use of Verificarlo to implement stochastic arithmetic through the Fuzzy
    environment enables to study of the impact of numerical perturbations
    introduced in non-Python dependencies while preserving the high-level view
    of Python variables and functions. For instance, the SciPy environment
    wraps many computation kernels written in C, such as BLAS or LAPACK, which
    have a considerable impact on the stability of applications using them.}

@@

% \gkmod{}{Note: this paragraph just reads like a list with no real
% interpretation/commentary. I think it would be better to state the strengths
% of this technique, cite examples, and then state limitations/why it isn't the
% only technique used.}
The source-to-source technique requires a rewriting of the application to modify
floating-point types and provides fine-grained control of the analyzed code.
CADNA~\cite{jezequel2008cadna} is a library for C, C++, and Fortran implementing
CESTAC~\cite{vignes1993stochastic} stochastic arithmetic.
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model
to propagate numerical errors. MCAlib~\cite{frechtling2015mcalib} is a C library
that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using
the MPFR~\cite{fousse2007mpfr} library. \add{ADAPT \mbox{~\cite{menon2018adapt}} uses
    Automatic Differentiation to compute a sensitivity profile for each variable in a
    program}. Finally, the work in~\cite{tang2016software} proposed a
source-to-source framework for executing targeted code in infinite and fixed
precision with and without stochastic arithmetic. The main drawback of the
source-to-source approach is its lack of scalability since rewriting large
codes can be a tedious task.

@@

The compiler-based approach instruments floating-point expressions at compile
time and so allows the user to automatically instrument large codebases.
Verificarlo~\cite{verificarlo} is a compiler that supports different types of
arithmetic instrumentations, including MCA. The work in~\cite{bao2013fly}
modified the GNU Compiler Collection (GCC) to track local floating-point errors
across executions. pLiner~\cite{guo2020pliner} is a root-cause analyzer based on
the Clang compiler that detects floating-point operations responsible for result
variability using a source-to-source transformation at the Abstract Syntax Tree
(AST) level to rewrite parts of code with higher precision.
PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel} \add{and
    NSan~\mbox{\cite{courbet2021nsan}} are compiler-based approaches}
that use shadow execution to detect numerical issues using higher
precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability
induced by compilers and their optimizations. The work
in~\cite{wang2012development} proposes a numerical debugger based on
GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on
FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and
Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB,
respectively. The main limitation of the compiler-based approach is that one
must have access to the source code.

>>>>

\ifANNOTATED
\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\DeclareRobustCommand{\add}[1]{\textcolor{ao(english)}{#1}}%\sethlcolor{lightgreen}\hl{#1}}
\DeclareRobustCommand{\remove}[1]{}
\DeclareRobustCommand{\addCite}[1]{\textcolor{lightgreen}{\mbox{#1}}}
\newcommand{\TG}[1]{\color{red}\textsc{From Tristan}: #1\xspace\color{black}}
\else
\DeclareRobustCommand{\add}[1]{#1}
\DeclareRobustCommand{\remove}[1]{}
\DeclareRobustCommand{\addCite}[1]{#1}
\newcommand{\TG}[1]{}
\fi

@@

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (\add{a}\remove{1}) generating numerically-perturbed
executions, (\add{b}\remove{2}) tracing floating-point computations along with
program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. \add{PyTracer addresses this work (a) by relying on
    Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
    computing libraries instrumented with Monte-Carlo arithmetic through the
    Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
    Python functions, modules, and classes through meta-programming, and (c) by
    visualizing stability measures throughout program executions with an
    interactive Plotly dashboard. \mbox{\pytracer} is available at
    \mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements
    with (1) a Fuzzy Python interpreter instrumented with Monte-Carlo
    arithmetic, (2) a dynamic instrumentation of Python functions, modules, and
    classes through meta-programming, (3) an interactive Plotly dashboard to
    visualize stability measures throughout program executions. }

@@

\add{
    To detect numerical instabilities, \pytracer works in three main steps:
    \begin{enumerate}
        \item It instruments Python programs and their module dependencies
              automatically, to save the input and output values of functions in a trace
              file.
        \item It executes the instrumented program several times using Monte-Carlo
              Arithmetic to inject perturbations in function outputs.
        \item It aggregates all trace files to compute statistics about the traced
              values, and visualize them in a GUI.
    \end{enumerate}
    \pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
    In Step 1, \pytracer uses monkey patching to replace the functions called
    by the application with a wrapper that saves their arguments and returned
    values in a trace file. In Step 2, it executes the application multiple
    times with Monte-Carlo Arithmetic as implemented in the
    Fuzzy~\cite{kiar2020comparing} environment. It then
    aggregates the resulting trace files to compute summary statistics on
    traced variables. In Step 3, a Plotly dashboard provides interactive
    visualizations highlighting unstable code sections.
}

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis~\cite{helton2006survey}, and random
seed analysis~\cite{hellekalek1998good}. \add{PyTracer is agnostic to the noise model used and can therefore
    be used for all these methods}.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis\add{.}

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
    simulates roundoff errors by applying function $inexact$ only to the
    output returned by a function. The Precision Bounding (PB) mode that applies
    the $inexact$ function on inputs can
    identify other types of instabilities, in particular catastrophic
    cancellations, however, it is more difficult to apply to large code bases
    as it often leads to execution errors that require human intervention.}

@@

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate the
    distribution will be. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample sizes (3 to 30)
    can give good estimates, the issue is to capture ``rare events" that could make the
    computation strongly diverge from the mean. Determining the number of
    MCA samples to use for a given instrumented code is still an open research question. 
%%     Recent theoretical works}~\cite{sohier2021confidence}
%% \add{on the estimation of significant digits allow computing an associated
%%   confidence interval depending on the sample size and the underlying distribution.
}

@@

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
    PyTracer. We observe a significant loss of \add{result} precision in some regions, which follows
    an interesting pattern that could be further investigated.
    This example shows that \pytracer workds on 2D inputs and the importance of
    visualizing complex data and not only reporting the values.}

@@

Replacing \texttt{gesdd} with\texttt{gesvd} solved the SVD convergence issue.
\pytracer allows us to quickly visualize the new \add{result} precision.
Figure~\ref{fig:face_recognition_svd} shows the \add{result} precision of the eigenvalues
computed with \texttt{gesvd} \add{in single precision}. \add{The result precision}
varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. \add{We can note that even with the \texttt{gesvd} method,
    the \add{result} precision of the singular values and eigenvectors is not uniform. 
    Indeed, we can see that the \add{result} precision of the lowest singular values and eigenvectors decreases 
    until reaching 0 significant bits.}

@@

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues such as
\href{https://github.com/google/TensorNetwork/pull/962}{github.com/google/TensorNetwork/pull/962} or
\href{https://github.com/scipy/scipy/issues/15024}{github.com/scipy/scipy/issues/15024}
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request at \href{https://github.com/scikit-learn/scikit-learn/pull/20617}{github.com/scikit-learn/scikit-learn/pull/20617}
has recently been submitted in the scikit-learn repository to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unified view of all traces allows the user to quickly navigate through the
    \add{result} precision of different variables.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}

@@

\add{\pytracer can be used with different types of numerical perturbations,
    including random seeds, data noise, or stochastic arithmetic. The
    use of Verificarlo to implement stochastic arithmetic through the Fuzzy
    environment enables us to study the impact of numerical perturbations
    introduced in non-Python dependencies while preserving the high-level view
    of Python variables and functions. For instance, the SciPy environment
    wraps many computation kernels written in C, such as BLAS or LAPACK, which
    have a considerable impact on the stability of applications using them.}

@@

% \gkmod{}{Note: this paragraph just reads like a list with no real
% interpretation/commentary. I think it would be better to state the strengths
% of this technique, cite examples, and then state limitations/why it isn't the
% only technique used.}
The source-to-source technique requires a rewriting of the application to modify
floating-point types and provides fine-grained control of the analyzed code.
CADNA~\cite{jezequel2008cadna} is a library for C, C++, and Fortran implementing
CESTAC~\cite{vignes1993stochastic} stochastic arithmetic.
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model
to propagate numerical errors. MCAlib~\cite{frechtling2015mcalib} is a C library
that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using
the MPFR~\cite{fousse2007mpfr} library. \add{ADAPT \mbox{~\cite{menon2018adapt}} uses
    Automatic Differentiation to compute a sensitivity profile for each variable in a
    program}. Finally, the work in~\cite{tang2016software} proposes a
source-to-source framework for executing targeted code in infinite and fixed
precision with and without stochastic arithmetic. The main drawback of the
source-to-source approach is its lack of scalability since rewriting large
codes can be a tedious task.

@@

The compiler-based approach instruments floating-point expressions at compile
time and so allows the user to automatically instrument large codebases.
Verificarlo~\cite{verificarlo} is a compiler that supports different types of
arithmetic instrumentations, including MCA. The work in~\cite{bao2013fly}
modified the GNU Compiler Collection (GCC) to track local floating-point errors
across executions. pLiner~\cite{guo2020pliner} is a root-cause analyzer based on
the Clang compiler that detects floating-point operations responsible for result
variability using a source-to-source transformation at the Abstract Syntax Tree
(AST) level to rewrite parts of code with higher precision.
FPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel} \add{and
    NSan~\mbox{\cite{courbet2021nsan}} are compiler-based approaches}
that use shadow execution to detect numerical issues using higher
precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability
induced by compilers and their optimizations. The work
in~\cite{wang2012development} proposes a numerical debugger based on
GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on
FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and
Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB,
respectively. The main limitation of the compiler-based approach is that one
must have access to the source code.
